Challenges and Solutions of Developing and Implementing a Novel Desktop-as-a-Service

### 2024-06-01 paper.tex
Final typo cheker commit
Cleaned up some sources and fixed some horizontal spacings between footnotes and text

For every deployment scenario or cloud service category possible,
users expect a latency value that is not annoying
when interacting with the service and its applications.
Choy et al.~\cite{ChoyWongSimonRosenberg2012},
Claypool et al.~\cite{claypool2010latency},
and Jarschel et al.~\cite{jarschel2011evaluation}
consider a maximum tolerable latency, based on subjective tests,
being around 100\,ms.
This relatively low value sets the demand for a fast network interaction
between client and service and limits their maximum distance.
As mentioned in Clincy et al.~\cite{clincy2013subjective},
The former cloud gaming service provider OnLive specified in the year 2010
that the distance between the service provider and the consumer
should not be greater than 1000 miles (approx. 1600\,km) because otherwise,
the round-trip communications delay time between will be too long for video games.
A more in-depth comparison of
commercially available cloud gaming services is given
by Di Domenico et al.~\cite{di2021network}
or by Carrascosa et al.~\cite{carrascosa2022cloud}.
Additionally, implied delays and round-trips can be further reduced
by using more sophisticated transmission protocols
directly interacting with GPU drivers,
such as Moonlight~\cite{hanimplementation}~\cite{darcy2022latency}.

@@

Guacamole supports redirecting audio, printing,
and disk access via the protocol RDP.
The printing feature does not allow access to local or remote printers;
instead, it allows users to print directly to PDF files,
which are received by the user's web browser.
As a prerequisite, the GhostScript interpreter
must be deployed on the Guacamole server.
File transfer over RDP supports Guacamole
by emulating a virtual disk drive that persistently stores data
on the Guacamole server.~\cite{ApacheGuacamole2024}

@@

Regarding Infrastructure-as-a-Service (IaaS),
our solution generally offers the flexibility
to use either virtual machines or containers.
Scientific publications discussing relevant performance aspects of such systems
highlight the apparent advantage of a lightweight container solution
over virtual machines when identical programs are tested in both environments
~\cite{felter2015updated,potdar2020performance}.

@@

Regarding Platform-as-a-Service (PaaS),
studies show that there is no dramatic difference in performance
when derivatives of operating systems are compared to each other
~\cite{balen2020performance,boras2020performance}.
Nevertheless, some derivatives are more lightweight and performant.
In contrast, comparing operating systems of different types
reveals more obvious results, particularly when considering
low-level operating system functionality implementations
such as interrupt handling, memory allocation, and boot time
~\cite{sulaiman2021comparison}.
Additional performance potentials can be leveraged
in cases where a Windows virtual machine can be replaced
with a container-based solution integrating wine
~\cite{huang2012performance}.

@@

Allthough we can not yet provide a full-fledged performance analysis,
we already measured promising results in our local test environment.
Depending on the exact setup,
the solution provides
reasonable round-trip times and delays for each communication direction
visualized in figure~\ref{perf_internal_communication}.
In a minimal setup, we measured delays of around 10ms for client communication,
around 100ms for internal backend communication
but up to 1000ms for instance communication.
But these numbers are only rough estimates
as certain optimization concepts remain to be implemented.
% Explanation delays: Half-way of a Roundtrip.
% We state 1000ms for instance-communication (500ms per connection, we make two, so 1000ms)
% We state 100ms for internal comm, which is a roundtrip of 200ms for auth divided by two, so 100ms
% 10ms for client delay we can take a clientside timestamp when sending request and a server side timestamp when the request is intially received.
% So we directly measure a deley, no rtt. The delay varias between 9 and 14ms, so we state roughly 10ms.
% In my newest tests this number goes up to 50ms, but thats ok for what we receive in advance.

@@

\subsection{Stability}
Stability in DaaS applications is especially important
for meeting the demands of continuous and time-critical operations
across diverse user scenarios.
Ensuring the stability of all system components during runtime is paramount
to maintain service continuity, handle growing user loads,
and transparent failure recovery while providing a seamless end-user experience.
These considerations are integral to our architectural concept,
directly impacting the operational integrity of the system
and its suitability for enterprise-level deployment and trust.

@@

\subsubsection{Scalability of Components}
Scalability in DaaS environments
can be approached through two fundamental methods: horizontal and vertical scaling.
Horizontal scaling (scale out), involves adding more machines or instances
to manage increased loads.
In contrast, vertical scaling (scale up),
adds more power (CPU, RAM) to existing machines
~\cite{vaquero2011dynamically}.
While horizontal scaling offers flexibility
and is often preferred in cloud environments,
vertical scaling remains important for applications
requiring strong single-thread performance or low latencies.

@@

	      and can be externalized to a dedicated systems.

@@

	      such as data replication, clustering, failover mechanisms or RAID control.

@@

This implies that any action triggered, whether in the frontend on the client side,
in the backend on the host side, or within hosted instances,
must ensure that all necessary information is readily available at the time of processing.

@@

If that is the case, each distributed host system might only contain it's essential components,
including the backend service with all hosted instances
as well as the described proxy component for direct user interaction.
Any other component may reside somewhere in the DaaS network.

@@

For Linux operating system deployments,
the free software projects xrdp and x11vnc enable all desired features
and single-app functionality can be easily obtained
by using a custom window manager such,
as for example, qtile~\footnote{\url{https://qtile.org/}}.

@@

One challenge was imposed by Guacamole's current lack of support
for dynamic desktop resizing with VNC,
leading to misaligned layouts and a poor user experience.
To address this until such features are available in
Guacamole~\footnote{\url{https://issues.apache.org/jira/browse/GUACAMOLE-1196}},
we implemented a separate mechanism using a WebSocket proxy
and direct interaction with each instance.

@@

For future work,
we plan to replace our temporary solution for internal communication
with a more efficient approach as our preliminary performance tests revealed
a potential bottlenack in that particular area.
We also plan to apply certain optimization strategies
with respect to backend and instance communication.
We further plan to enhance our yet limited performance testing
and will provide a full-fledged performance analysis
of realistic deployment scenarios and utilization patterns
which include an evaluation of different user behaviors
within various popular applications.

>>>>

For every deployment scenario or cloud service category possible,
users expect a latency value that is not annoying
when interacting with the service and its applications.
Choy et al.~\cite{ChoyWongSimonRosenberg2012},
Claypool et al.~\cite{claypool2010latency},
and Jarschel et al.~\cite{jarschel2011evaluation}
consider a maximum tolerable latency, based on subjective tests,
being around 100\,ms.
This relatively low value sets the demand for a fast network interaction
between client and service and limits their maximum distance.
As mentioned in Clincy et al.~\cite{clincy2013subjective},
The former cloud gaming service provider OnLive specified in the year 2010
that the distance between the service provider and the consumer
should not be greater than 1000 miles (approx. 1600\,km) because otherwise,
the round-trip communications delay time between will be too long for video games.
A more in-depth comparison of
commercially available cloud gaming services is given
by Di Domenico et al.~\cite{di2021network}
or by Carrascosa et al.~\cite{carrascosa2022cloud}.
Additionally, implied delays and round-trips can be further reduced
by using more sophisticated transmission protocols
directly interacting with GPU drivers,
such as Moonlight~\cite{darcy2022latency,hanimplementation}.

@@

Guacamole supports redirecting audio, printing,
and disk access via the protocol RDP.
The printing feature does not allow access to local or remote printers;
instead, it allows users to print directly to PDF files,
which are received by the user's web browser.
As a prerequisite, the GhostScript interpreter
must be deployed on the Guacamole server.
File transfer over RDP supports Guacamole
by emulating a virtual disk drive that persistently stores data
on the Guacamole server.\footnote{\url{https://guacamole.apache.org/doc/1.5.5/gug/configuring-guacamole.html}}

@@

Regarding Infrastructure-as-a-Service (IaaS),
our solution generally offers the flexibility
to use either virtual machines or containers.
Scientific publications discussing relevant performance aspects of such systems
highlight the apparent advantage of a lightweight container solution
over virtual machines when identical programs are tested in both
environments~\cite{felter2015updated,potdar2020performance}.

@@

Regarding Platform-as-a-Service (PaaS),
studies show that there is no dramatic difference in performance
when derivatives of operating systems are compared to each
other~\cite{balen2020performance,boras2020performance}.
Nevertheless, some derivatives are more lightweight and performant.
In contrast, comparing operating systems of different types
reveals more obvious results, particularly when considering
low-level operating system functionality implementations
such as interrupt handling, memory allocation, and boot
time~\cite{sulaiman2021comparison}.
Additional performance potentials can be leveraged
in cases where a Windows virtual machine can be replaced
with a container-based solution integrating 
wine~\cite{huang2012performance}.

@@

Although we can not yet provide a full-fledged performance analysis,
we already measured promising results in our local test environment.
Depending on the exact setup,
the solution provides
reasonable round-trip times and delays for each communication direction
visualized in figure~\ref{perf_internal_communication}.
In a minimal setup, we measured delays of around 10ms for client communication,
around 100ms for internal backend communication
but up to 1000ms for instance communication.
However, these numbers are only rough estimates
as certain optimization concepts remain to be implemented.
% Explanation delays: Half-way of a Roundtrip.
% We state 1000ms for instance-communication (500ms per connection, we make two, so 1000ms)
% We state 100ms for internal comm, which is a roundtrip of 200ms for auth divided by two, so 100ms
% 10ms for client delay we can take a clientside timestamp when sending request and a server side timestamp when the request is intially received.
% So we directly measure a deley, no rtt. The delay varias between 9 and 14ms, so we state roughly 10ms.
% In my newest tests this number goes up to 50ms, but thats ok for what we receive in advance.

@@

\subsection{Stability}
Stability in DaaS applications is essential
for meeting the demands of continuous and time-critical operations
across diverse user scenarios.
Ensuring the stability of all system components during runtime is paramount
to maintain service continuity, handle growing user loads,
and transparent failure recovery while providing a seamless end-user experience.
These considerations are integral to our architectural concept,
directly impacting the operational integrity of the system
and its suitability for enterprise-level deployment and trust.

@@

\subsubsection{Scalability of Components}
Scalability in DaaS environments
can be approached through two fundamental methods: horizontal and vertical scaling.
Horizontal scaling (scale out), involves adding more machines or instances
to manage increased loads.
In contrast, vertical scaling (scale up),
adds more power (CPU, RAM) to existing 
machines~\cite{vaquero2011dynamically}.
While horizontal scaling offers flexibility
and is often preferred in cloud environments,
vertical scaling remains important for applications
requiring strong single-thread performance or low latencies.

@@

	      and can be externalized to a dedicated system.

@@

	      such as data replication, clustering, failover mechanisms, or RAID control.

@@

This implies that any action triggered, whether in the frontend on the client side,
in the backend on the host side or within hosted instances,
must ensure that all necessary information is readily available at the time of processing.

@@

If that is the case, each distributed host system might only contain its essential components,
including the backend service with all hosted instances
as well as the described proxy component for direct user interaction.
Any other component may reside somewhere in the DaaS network.

@@

For Linux operating system deployments,
the free software projects xrdp and x11vnc enable all desired features
and single-app functionality can be easily obtained
by using a custom window manager such,
as for example, qtile\footnote{\url{https://qtile.org}}.

@@

One challenge was imposed by Guacamole's current lack of support
for dynamic desktop resizing with VNC,
leading to misaligned layouts and a poor user experience.
To address this until such features are available in
Guacamole\footnote{\url{https://issues.apache.org/jira/browse/GUACAMOLE-1196}},
we implemented a separate mechanism using a WebSocket proxy
and direct interaction with each instance.

@@

For future work,
we plan to replace our temporary solution for internal communication
with a more efficient approach as our preliminary performance tests revealed
a potential bottleneck in that particular area.
We also plan to apply certain optimization strategies
concerning backend and instance communication.
We further plan to enhance our yet limited performance testing
and will provide a full-fledged performance analysis
of realistic deployment scenarios and utilization patterns
which include an evaluation of different user behaviors
within various popular applications.
